---
title: "Tibet Pollen"
output: html_notebook
---

```{r setup}
library(tidyverse)
library(readxl)
library(sf)
library(scico)
library(stars)

china <- rnaturalearth::ne_countries(country = 'China', returnclass = 'sf')
```

```{r}
modern <- read_excel('~/Downloads/pollen sites modern China - old+new collection in press.xlsx') %>%
  st_as_sf(coords = c(4,5))

fossil <- read_excel('~/Downloads/pollen sites fossil China - Ni 2014 palaeo3.xlsx') %>%
  st_as_sf(coords = c(8,9))


fossil2 <- read_excel('~/Downloads/china-fossil-pollen-from Ni 2014.xlsx') %>%
  st_as_sf(coords = c(3,4))

fossil1
fossil2
fossil3 %>% count(Site)
fossil4 %>% count(`Site/Taxa`)
```

```{r}
plot(modern)
```

```{r}
plot(fossil)
```

```{r}
head(modern)
```

```{r}
count(modern, `Sample type`) %>%
ggplot(aes(reorder(`Sample type`, n), n)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(title = 'Modern pollen sample distribution', y = 'Count', x = 'Sample type')
```

```{r}
count(modern, `Data source`) %>%
ggplot(aes(reorder(`Data source`, n), n)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(title = 'Modern pollen data sources', y = 'Count', x = 'Data source')
```

```{r}
count(modern, `Data type`) %>%
ggplot(aes(reorder(`Data type`, n), n)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(title = 'Modern pollen data type', y = 'Count', x = 'Data type')
```

```{r}
count(modern, `Modern vegetation information (sample and its surroundings)`) %>%
ggplot(aes(reorder(`Modern vegetation information (sample and its surroundings)`, n), n)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(title = 'Modern pollen data type', y = 'Count', x = 'Data type')
```

```{r}
modern
```


```{r}
pollen <- read_excel('china pollen data/China modern pollen - data.xlsx', sheet = 2, skip = 1) %>%
  slice(-(1:2)) %>%
  select(-(`No.`:`Elevation (m)`)) %>%
  mutate(across(everything(), as.numeric))
```

```{r}
pollen_fossil <- read_excel('china pollen data/China west fossil pollen - data.xlsx') %>%
  slice(-(1:2)) %>%
  column_to_rownames("Site/Taxa") %>%
  select(-(Long:Age)) %>%
  mutate(across(everything(), as.numeric))

# new fossil pollen data has some duplicated sample names with different values
pollen_fossil_new <- read_excel('~/Downloads/China west fossil pollen - data from sites.xlsx') %>%
  select(-(Site:Age))

fossil_loc <- read_excel('china pollen data/China west fossil pollen - data.xlsx') %>%
  slice(-(1:2)) %>%
  select(`Site/Taxa`:Age) %>%
  separate(`Site/Taxa`, into = c('site', 'year'), sep = '_', extra = 'merge') %>%
  # this part required because there are sometimes extra '_'  in the site names
  separate(year, into = c('year1', 'year2'), sep = '_') %>% # this warning is ok
  mutate(site = if_else(is.na(year2), site, paste(site, year1, sep = '_'))) %>%
  select(-year1, -year2)

# fossil_loc and fossil_loc2 are the same except for one or three samples that are missing. both have some unsual spatial locations
fossil_loc_new <- read_excel('~/Downloads/China west fossil pollen - data from sites.xlsx') %>%
  select(Site:Age) %>%
  separate(Site, into = c('site', 'year'), sep = '_', extra = 'merge') %>%
  # this part required because there are sometimes extra '_'  in the site names
  separate(year, into = c('year1', 'year2'), sep = '_') %>% # this warning is ok
  mutate(site = if_else(is.na(year2), site, paste(site, year1, sep = '_'))) %>%
  select(-year1, -year2)

fossil_loc %>% count(site) 
fossil_loc_new %>% count(site) 

ggplot(fossil_loc_new)  +
  geom_sf(data= china)+ geom_point(aes(Long, Lat)) +
  geom_sf(data = st_set_crs(fossil, 4326), color = 'red') +
  geom_sf(data = st_set_crs(fossil2, 4326), color = 'green')


read_excel('~/Downloads/China west fossil pollen - data from sites.xlsx') %>% filter(Site == 'SG1_0737900')
```


Check that all rows sum to 100%
```{r}
all.equal(sum(rowSums(pollen)) / 100,nrow(pollen))
rowSums(pollen_fossil) %>% hist
rowSums(pollen_fossil_new) %>% near(100, tol = .01) %>% .[. == FALSE]

pollen_fossil_new %>% mutate(sums = rowSums(.)) %>%
  filter(sums < 99) %>%
  select(sums)

```


```{r}
pca <- prcomp(pollen, scale. = FALSE)
```

```{r}
screeplot(pca, type = 'lines', npcs = 20)
```

```{r}
pca$rotation[,1:2] %>% plot
```

so what's up with these taxa that are pca1 scaled?
```{r}
pca$rotation[,1:2] %>% as.data.frame %>% filter(PC1 > .2 | PC1 < -.2)
```

Betula is birch, fraxinus is ash, this is all arborieal, day lily (rare)
```{r}
pca_s$rotation[,1:2] %>% as.data.frame %>% filter(PC1 > .2)
```

sedges are extreme on pc2 unscaled, grow along bodies of water, major grassland biomes
```{r}
pca$rotation[,1:2] %>% as.data.frame %>% filter(PC2 > .2)
```
```{r}
pca_s$rotation[,1:2] %>% as.data.frame %>% filter(PC2 < -.2)
```
# Climate relationships

So let's build a model predict MAT and TAP from the leading 2 PCs (arborial-nonarboreal and sedges)

```{r}
pc_scores <- pca$x %>% 
  as_tibble() %>%
  bind_cols(modern, .) %>%
  select(PC1, PC2) %>%
  st_set_crs(4326)
```


```{r}
ggplot() +
    geom_sf(data= china) +
  geom_sf(data = pc_scores, aes(color = PC1)) + 
  scale_color_scico(palette = 'bamako', limits = c(-70, 70), direction = -1) +
  theme_bw() +
  ggtitle('Modern Arboreal/Non-Arboreal Pollen concentration')
```
```{r}
ggplot() +
    geom_sf(data= china) +
  geom_sf(data = pc_scores, aes(color = PC2)) + 
  scale_color_scico(palette = 'bamako', limits = c(-80, 80), direction = -1) +
  theme_bw() +
  ggtitle('Modern Sedge/Non-Sedge Pollen concentration')
```

```{r}
clim <- c('/Volumes/Data/CHELSA/CHELSA_bio1_1981-2010_V.2.1.tif',
  '/Volumes/Data/CHELSA/CHELSA_bio12_1981-2010_V.2.1.tif',
  '/Volumes/Data/CHELSA/CHELSA_gdd10_1981-2010_V.2.1.tif',
  '/Volumes/Data/CHELSA/CHELSA_npp_1981-2010_V.2.1.tif',
  '/Volumes/Data/CHELSA/CHELSA_gdd0_1981-2010_V.2.1.tif') %>%
  read_stars() %>%
  st_crop(st_bbox(pc_scores)) %>%
  setNames(c('MAT', 'TAP', 'GDD10', 'NPP', 'GDD0')) %>%
  mutate(GDD10 = replace_na(GDD10, 0),
         GDD0 = replace_na(GDD0, 0)) %>%
  st_as_stars()
```

```{r}
plot(clim)
plot(clim['TAP'])
plot(clim['GDD10'])
plot(clim['GDD0'])
plot(clim['NPP'])
```


```{r}
clim %>%
  st_as_stars %>%
  as_tibble %>%
  ggplot(aes(MAT, GDD10)) +
  geom_point()
```


```{r}
pts <- st_extract(clim, at = pc_scores) # bilinear = TRUE causes this to fail for some reason
```

```{r}
library(mgcv)
dat <- pc_scores %>%
  mutate(mat = pts$MAT,
         tap = pts$TAP,
         gdd10 = pts$GDD10,
         gdd0 = pts$GDD0,
         npp = pts$NPP,
         x = st_coordinates(.)[,1],
         y = st_coordinates(.)[,2])
```


```{r}
m1 <- gam(mat ~ s(PC1, bs = 'cr') + s(PC2, bs = 'cr'), method = 'REML', select = TRUE, data = dat)
m2 <- gam(tap ~ s(PC1, bs = 'cr') + s(PC2, bs = 'cr'), method = 'REML', select = TRUE, data = dat)
m3 <- gam(gdd10 ~ s(PC1, bs = 'cr') + s(PC2, bs = 'cr'), method = 'REML', select = TRUE, data = dat)
m4 <- gam(npp ~ s(PC1, bs = 'cr') + s(PC2, bs = 'cr'), method = 'REML', select = TRUE, data = dat)
m5 <- gam(gdd0 ~ s(PC1, bs = 'cr') + s(PC2, bs = 'cr'), method = 'REML', select = TRUE, data = dat)

summary(m1)$r.sq
summary(m2)$r.sq
summary(m3)$r.sq
summary(m4)$r.sq
summary(m5)$r.sq
```

```{r}
plot(m1)
```

```{r}
plot(m2)
```

```{r}
plot(m3)
```

```{r}
plot(m4)
```

```{r}
pc_scores %>%
  ggplot(aes(PC1, PC2)) +
  geom_point(alpha = .2)
```
How about the forward model?
```{r}
m5 <- gam(PC1 ~ s(mat, bs = 'cr') + s(npp, bs = 'cr') + s(tap, bs = 'cr') + s(gdd10, bs = 'cr'), data = dat, method = 'REML', select = TRUE)

summary(m5)$r.sq
```
```{r}
plot(m5, scale = 0)
```
## Modern Analogue technique

```{r}
library(analogue)
```

```{r}
plot(pts)
```

```{r}
dat <- analogue::join(pollen, pollen_fossil, verbose = TRUE, type = 'inner')
```


```{r}
gdd_mat <- mat(dat$pollen / 100, as_tibble(pts)$GDD10, method = "SQchord")
```

```{r}
gdd_mat
```

```{r}
summary(gdd_mat)
```
```{r}
getK(gdd_mat)
```

```{r}
plot(gdd_mat)
```

```{r}
gdd_boot <- bootstrap(gdd_mat, n.boot = 25)
gdd_boot
#setK(gdd_mat) <- getK(gdd_boot)
```

```{r}
plot(pts %>% select(GDD10))
```

```{r}
pastgdd <- predict(gdd_mat, dat$pollen_fossil / 100)
pastgdd$predictions$model$predicted %>% colMeans()
```

```{r}
pastgdd
```

```{r}
dat$pollen_fossil %>% as_tibble() %>% mutate(test = pastgdd)
```

```{r}

dat$pollen_fossil
as_tibble(pts)$GDD10
```


```{r}
library(rioja)
rioja::MAT()
gdd_mat <- rioja::MAT(dat$pollen / 100, as_tibble(pts)$GDD10, )
performance(gdd_mat)
```

```{r}
print(gdd_mat)
```

```{r}
screeplot(gdd_mat)
```

```{r}
pastgdd <- predict(gdd_mat, dat$pollen_fossil / 100, k = 4)

fossil_loc <- read_excel('china pollen data/China west fossil pollen - data.xlsx') %>%
  slice(-(1:2)) %>%
  select(`Site/Taxa`:Age)

recon <- as_tibble(pastgdd$fit, rownames = 'Site/Taxa') %>%
  left_join(fossil_loc) %>%
  separate(`Site/Taxa`, into = c('site', 'year'), sep = '_', extra = 'merge') %>%
  # this part required because there are sometimes extra '_'  in the site names
  separate(year, into = c('year1', 'year2'), sep = '_') %>% # this warning is ok
  mutate(site = if_else(is.na(year2), site, paste(site, year1, sep = '_'))) %>%
  select(-year1, -year2, -MAT) %>%
  rename(gdd10 = MAT.wm)
```
```{r}
recon %>% count(site)
```
```{r}
plot(fossil)
```


```{r}

test <- bam(gdd10 ~ te(Long, Lat, Age, k = c(50, 20), d = c(2, 1), bs = c('tp', 'cr')), data = recon %>% filter(Long > 60, Age < 12000))


hist(residuals.gam(test))
summary(test)
gam.check(test)
hist(recon$Long) ;hist(recon$Lat)

ggplot(recon, aes(Long, Lat)) + geom_point()
```
```{r}
recon;fossil
```


```{r}
plot(test)
```
```{r}
recon %>% filter(Age < 12000) %>% pull(Age) %>% hist
```

```{r}
china_hex <- china %>% st_make_grid(n = 50, square = FALSE) %>% st_as_sf() %>% mutate(Long = st_coordinates(st_centroid(.))[,1],
                                                           Lat = st_coordinates(st_centroid(.))[,2]) %>%
    filter(st_intersects(x, china, sparse = FALSE)) %>%
  mutate(id = 1:n())
   
map_dfr(c(0, 2000, 4000, 6000), ~ mutate(china_hex, Age = .)) %>%
    mutate(pred = predict(test, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  ggplot() + geom_sf(aes(fill = pred) ,size = .1) + scale_fill_viridis_c(option = 'magma') +
    facet_wrap(~Age) +
  theme_bw() 
```

```{r}
mutate(china_hex, Age = 0) %>%
    mutate(pred = predict(test, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  mutate(Age = 6000) %>%
      mutate(pred2 = predict(test, ., type = 'response')) %>%
  mutate(pred2 = if_else(pred2 < 0, 0, pred2)) %>%
  mutate(pred = pred - pred2) %>%
  ggplot() + geom_sf(aes(fill = pred) ,size = .1) + scale_fill_distiller(palette ='RdBu', limits = c(-4000, 4000)) +
  theme_bw() 
```
```{r}
map_dfr(seq(12000, 0, -1000), ~ mutate(china_hex, Age = .)) %>%
    mutate(pred = predict(test, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  ggplot() + geom_sf(aes(fill = pred > 2900) ,size = .1) + scale_fill_viridis_d(option = 'magma') +
    facet_wrap(~Age) +
  theme_bw() 
```

```{r}
library(exactextractr)
gdd_dgg <- exact_extract(as(clim['GDD10'], 'Raster'), china_hex, 'mean')

china_hex %>% transmute(gdd_dgg) %>% ggplot() + geom_sf(aes(fill = gdd_dgg), size = .1) + scale_fill_viridis_c(option = 'magma') + theme_bw()
```
```{r}
china_hex %>% transmute(gdd_dgg) %>% ggplot() + geom_sf(aes(fill = gdd_dgg > 2900), size = .1) + scale_fill_viridis_d(option = 'magma') + theme_bw()
```
repeat with bins
```{r}

pollen;modern
gdd_mat <- rioja::MAT(dat$pollen / 100, as_tibble(pts)$GDD10, )
pastgdd <- predict(gdd_mat, dat$pollen_fossil / 100, k = 4)

read_excel('china pollen data/China modern pollen - data.xlsx', sheet = 2, skip = 1) %>%
  slice(-(1:2)) %>%
  st_as_sf(coords = c(4,5), crs = 4326) %>%
 st_intersects(china_hex, .) %>%
  lengths %>%
  transmute(china_hex, .) %>% plot
```

```{r}
mod_hex_pol <- read_excel('china pollen data/China modern pollen - data.xlsx', sheet = 2, skip = 1) %>%
  slice(-(1:2)) %>%
  st_as_sf(coords = c(4,5), crs = 4326) %>%
  st_join(china_hex, join = st_within) %>% 
  select(Abelia:id) %>%
  st_drop_geometry() %>%
  mutate(across(everything(), as.numeric)) %>%
    group_by(id) %>%
  summarise(across(everything(), mean)) %>%
  left_join(select(china_hex, -Long, -Lat), .) %>%
  select(-Long, -Lat) %>% 
  st_drop_geometry() %>% as_tibble%>% select(-id) %>% 
  `/`(100)  



dat2 <- mod_hex_pol %>% mutate(gdd  = gdd_dgg) %>% remove_missing()

mat_bin <- MAT(dat2 %>% select(-gdd), dat2$gdd)
screeplot(mat_bin)
```


```{r}

fossil_loc_bin <- read_excel('china pollen data/China west fossil pollen - data.xlsx') %>%
  slice(-(1:2)) %>%
  st_as_sf(coords = c(2,3), crs = 4326) %>%
  st_join(china_hex, join = st_within) %>% 
  select(Age,`Abies+Picea`:id) %>%
  st_drop_geometry() %>%
  mutate(across(everything(), as.numeric)) %>%
  left_join(select(china_hex, -Long, -Lat), .) %>%
  select(-Long, -Lat)


fossil_poll_bin <- fossil_loc_bin %>% 
  st_drop_geometry() %>% as_tibble %>% remove_missing() %>% group_by(id, Age) %>% summarise(across(everything(), mean), .groups = 'drop') %>% mutate(no = interaction(id, Age, sep = '_')) %>% column_to_rownames(var = 'no') %>% 
  select(-id, -Age) %>%
  `/`(100)  

fossil_poll_bin_ages <- fossil_loc_bin %>% 
  st_drop_geometry() %>% as_tibble%>% select(id, Age) %>% remove_missing()

past_gdd_bin <- predict(mat_bin, fossil_poll_bin, k = 4)

recon_mod <- as_tibble(past_gdd_bin$fit, rownames = 'id') %>%
  separate(id, into = c('id', 'Age'), sep = '_') %>% 
  mutate(id = as.numeric(id), Age = as.numeric(Age)) %>%
 # left_join(fossil_poll_bin_ages) %>%
  select(id, Age, gdd10 = MAT.wm) %>%
  left_join(china_hex, by = 'id') %>%
  filter(Long > 60, Age < 10000) %>%
   bam(gdd10 ~ te(Long, Lat, Age, k = c(40, 30), d = c(2, 1), bs = c('tp', 'cr')), data = .)

summary(recon_mod)
plot(recon_mod)
```
```{r}

# not binned until end
mutate(china_hex, Age = 0) %>%
    mutate(pred = predict(test, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  mutate(Age = 6000) %>%
      mutate(pred2 = predict(test, ., type = 'response')) %>%
  mutate(pred2 = if_else(pred2 < 0, 0, pred2)) %>%
  mutate(pred = pred - pred2) %>%
  ggplot() + geom_sf(aes(fill = pred) ,size = .1) + scale_fill_distiller(palette ='RdBu', limits = c(-4000, 4000)) +
  theme_bw() 

#binned
mutate(china_hex, Age = 0) %>%
    mutate(pred = predict(recon_mod, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  mutate(Age = 6000) %>%
      mutate(pred2 = predict(recon_mod, ., type = 'response')) %>%
  mutate(pred2 = if_else(pred2 < 0, 0, pred2)) %>%
  mutate(pred = pred - pred2) %>%
  ggplot() + geom_sf(aes(fill = pred) ,size = .1) + scale_fill_distiller(palette ='RdBu', limits = c(-4000, 4000)) +
  theme_bw() 
```
```{r}
#why isn't there more variability here?
as_tibble(past_gdd_bin$fit, rownames = 'id') %>%
  separate(id, into = c('id', 'Age')) %>%
  mutate(id = as.numeric(id), Age = as.numeric(Age)) %>%
 # left_join(fossil_poll_bin_ages) %>%
  select(id, Age, gdd10 = MAT.wm) %>%
  filter(Age < 12000) %>%
  ggplot(aes(-Age, gdd10, color = as.factor(id))) + 
  geom_point(alpha = .5) +
  geom_line(alpha = .3) +
  theme_bw() +
  guides(color = 'none')
```

```{r}
data(columb.polys) ## district shapes list

spdep::poly2nb(china_hex )
columb.polys
```

```{r}
raster::getData()
```

