---
title: "Pollen-based reconstruction of growing degree days over Tibet"
output: html_notebook
---

# Setup and Preproccesing

Import the packages needed for this analysis. Also load a shapefile of China and the Tibetan Plateau to use in plots.
```{r setup}
library(tidyverse)
library(readxl)
library(sf)
library(scico)
library(stars)
library(rioja)
library(mgcv)
library(exactextractr)
library(fields)

china <- rnaturalearth::ne_countries(country = 'China', returnclass = 'sf')
tibet <- read_sf('~/Downloads/DBATP/DBATP_Polygon.shp') # source http://www.geodoi.ac.cn/weben/doi.aspx?Id=135
```

## Modern Pollen Data

Import the modern pollen data and site information.
```{r}
modern_sites <- read_excel('china pollen data/China modern pollen - info.xlsx', sheet = 2, skip = 1) %>%
  st_as_sf(coords = c(4,5), crs = 4326) %>%
  select(id = No., 
         site = `Site name`, 
         elev = `Elevation (m)`, 
         sample_type = `Sample type`, 
         source = `Data source`, 
         data_type = `Data type`, 
         modern_veg_map = `Modern vegetation information (the Vegetation Map of China)`, 
         modern_veg_samp = `Modern vegetation information (sample and its surroundings)`, 
         pot_veg = `Potential vegetation (for cultivated vegetation, aquatic vegetation and other cryptic vegetation)`)

modern_pollen <- read_excel('china pollen data/China modern pollen - data.xlsx', sheet = 2, skip = 1) %>%
  slice(-(1:2)) %>%
  select(-(Site:`Elevation (m)`)) %>%
  column_to_rownames(var = 'No.') %>%
  mutate(across(everything(), as.numeric))
```

Visualize the site data.
```{r echo=FALSE}
plot(modern_sites)
```

Plot some summaries of the site data.
```{r echo=FALSE}
count(modern_sites, sample_type) %>%
ggplot(aes(reorder(sample_type, n), n)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(title = 'Modern pollen sample distribution', y = 'Count', x = 'Sample type')
```
```{r}
ggplot(modern_sites) +
  geom_sf(data = china) +
  geom_sf(aes(color = sample_type)) +
  theme_bw()
```


Check that all rows sum to 100, and no single value is greater than 100
```{r}
all.equal(sum(rowSums(modern_pollen)) / 100, nrow(pollen))

sum(rowSums(modern_pollen > 100))
```

## Fossil Pollen Data

```{r}
fossil_sites <- read_excel('china pollen data/China west fossil pollen - data.xlsx') %>%
  slice(-(1:2)) %>%
  select(`Site/Taxa`:Age) %>%
  separate(`Site/Taxa`, into = c('site', 'year'), sep = '_', extra = 'merge') %>%
  # this part required because there are sometimes extra '_'  in the site names
  separate(year, into = c('year1', 'year2'), sep = '_') %>% # this warning is ok
  mutate(site = if_else(is.na(year2), site, paste(site, year1, sep = '_'))) %>%
  select(-year1, -year2) %>%
  # the CAOTANH site has lat and long coordinates reversed
  mutate(lon = if_else(site == 'CAOTANH', Lat, Long),
         lat = if_else(site == 'CAOTANH', Long, Lat)) %>%
  select(-Long, -Lat) %>%
  filter(site != 'PEARL14') %>%
  st_as_sf(coords = c(4,5), crs = 4326)

# these are different from the associated raw data -- specifically lacking that "transect"
fossil_sites1 <- read_excel('china pollen data/China west fossil pollen - info.xlsx', skip = 1) %>%
  # this has two lat lon fields, the unselected ones appear to be in degrees and meters?
  select(site = Site, elev = `Alt (m)`, type = `Sample Type`, lon = `Lon (d)`, lat = `Lat (d)`) %>%
  # Tanggula Mt has flipped lat lon coordinates
  mutate(lon2 = if_else(site == 'Tanggula Mt Co', lat, lon),
         lat2 = if_else(site == 'Tanggula Mt Co', lon, lat),
         lon = lon2, lat = lat2) %>%
  select(-lon2, -lat2) %>%
  st_as_sf(coords = c(4,5), crs = 4326)
    
# these are the same locations as china west fossil pollen - data
fossil_sites2 <- read_excel('~/Downloads/China west fossil pollen - data from sites.xlsx') %>%
  select(Site:Age) %>%
    slice(-n()) %>% # the last row of this file are column sums, so drop it
  separate(Site, into = c('site', 'year'), sep = '_', extra = 'merge') %>%
  # this part required because there are sometimes extra '_'  in the site names
  separate(year, into = c('year1', 'year2'), sep = '_') %>% # this warning is ok
  mutate(site = if_else(is.na(year2), site, paste(site, year1, sep = '_'))) %>%
  select(-year1, -year2) %>%
  # the CAOTANH site has lat and long coordinates reversed
  mutate(lon = if_else(site == 'CaotanH', Lat, Long),
         lat = if_else(site == 'CaotanH', Long, Lat)) %>%
  filter(site != 'Pearl14') %>%
  select(-Long, -Lat) %>%
  st_as_sf(coords = c(4,5), crs = 4326)
```

```{r}
ggplot(china)  +
  geom_sf(data= china) +
  geom_sf(data = fossil_sites, color = 'red', size = 2) +
  geom_sf(data = fossil_sites1, color = 'green')
```


```{r}
## these are the same sites, the former has the data. not all sites in the original file are in the pollen file from jade
fossil_sites3 <- read_excel('~/Downloads/china-fossil-pollen-from Ni 2014.xlsx') %>%
  st_as_sf(coords = c(3,4), crs = 4326)

fossil_sites4 <- read_excel('~/Downloads/pollen sites fossil China - Ni 2014 palaeo3.xlsx') %>%
  st_as_sf(coords = c(8,9), crs = 4326)

ggplot(china)  +
  geom_sf(data= china) +
    geom_sf(data = fossil_sites4, color = 'blue', size = 2) +
  geom_sf(data = fossil_sites3, color = 'red') 
```

```{r}
fossil_pollen <- read_excel('china pollen data/China west fossil pollen - data.xlsx') %>%
  slice(-(1:2)) %>%
  filter(!str_detect(`Site/Taxa`, 'PEARL')) %>% # the "transect"
  column_to_rownames("Site/Taxa") %>%
  select(-(Long:Age)) %>%
  mutate(across(everything(), as.numeric))

# new fossil pollen data has some duplicated sample names with different values
fossil_pollen2 <- read_excel('~/Downloads/China west fossil pollen - data from sites.xlsx') %>%
  slice(-n()) %>% # the last row of this file are column sums, so drop it
  filter(Site != 'NAQ22_-010', # the percentages from this sample only add to 5
         !str_detect(Site, 'Pearl')) %>% # the 'transect' 
  select(-Pinus_haploxylon) %>% # this taxon appears twice (with upper and lowercase H), delete this column as the total is much smaller
  select(-(Site:Age))

fossil_pollen3 <- read_excel('~/Downloads/china-fossil-pollen-from Ni 2014.xlsx') %>%
  select(-(`Site biomise`:Taxa))
```

Check that all rows sum to 100, and no single value is greater than 100
```{r}
all.equal(sum(rowSums(fossil_pollen)) / 100, nrow(fossil_pollen))
sum(rowSums(fossil_pollen > 100))

all.equal(sum(rowSums(fossil_pollen2)) / 100, nrow(fossil_pollen2))
sum(rowSums(fossil_pollen2 > 100))

all.equal(sum(rowSums(fossil_pollen3)) / 100, nrow(fossil_pollen3))
sum(rowSums(fossil_pollen3 > 100))

rowSums(fossil_pollen) %>% hist
rowSums(fossil_pollen2) %>% hist
rowSums(fossil_pollen3) %>% hist

dim(fossil_pollen)
dim(fossil_pollen2)
dim(fossil_pollen3)

setdiff(names(fossil_pollen), names(fossil_pollen2))
setdiff(names(fossil_pollen2), names(fossil_pollen))
```

# Modern Climate

## Modern Pollen 

Do a PCA on the modern pollen data to look at large scale pollen-climate relationships.

```{r}
pca <- prcomp(modern_pollen, scale. = FALSE)
pca_s <- prcomp(sqrt(modern_pollen), scale. = FALSE) # sqrt is hellinger transformation
```
```{r}
library(fastICA)
ica <- fastICA(modern_pollen, 9)
```

```{r}
screeplot(pca, npcs = 20)
screeplot(pca_s, type = 'lines', npcs = 20)
```
So PC1 is arboreal/non-arboreal. PC2 separates out sedges, grow along bodies of water, major grassland biomes
```{r}
pca$rotation[,1:2] %>% plot
pca$rotation[,1, drop = FALSE] %>% as.data.frame %>% filter(PC1 > .2 | PC1 < -.2)
pca$rotation[,2, drop = FALSE] %>% as.data.frame %>% filter(PC2 > .2 | PC2 < -.2)
```
```{r}
t(ica$A) %>% set
t(ica$A)[,1:2] %>% plot
t(ica$A)[,4:5] %>% plot

t(ica$A)[,1, drop = FALSE] %>% as.data.frame(row.names = names(modern_pollen)) %>% filter(V1 < -10)
t(ica$A)[,2, drop = FALSE] %>% as.data.frame(row.names = names(modern_pollen)) %>% filter(V1 > 10)
t(ica$A)[,3, drop = FALSE] %>% as.data.frame(row.names = names(modern_pollen)) %>% filter(V1 > 10)
t(ica$A)[,4, drop = FALSE] %>% as.data.frame(row.names = names(modern_pollen)) %>% filter(V1 > 10)
t(ica$A)[,5, drop = FALSE] %>% as.data.frame(row.names = names(modern_pollen)) %>% filter(V1 < -10)


```

Same patterns with the hellinger transformation
```{r}
pca_s$rotation[,1:2] %>% plot
pca_s$rotation[,1, drop = FALSE] %>% as.data.frame %>% filter(PC1 > .2 | PC1 < -.2)
pca_s$rotation[,2, drop = FALSE] %>% as.data.frame %>% filter(PC2 > .2 | PC2 < -.2)
```

### PCA


```{r}
pc_scores <- pca$x %>% 
  as_tibble() %>%
  bind_cols(modern_sites, .) %>%
  select(PC1:PC9) %>%
  st_set_crs(4326)
pc_scores_s <- pca_s$x %>% 
  as_tibble() %>%
  bind_cols(modern_sites, .) %>%
  select(PC1:PC9) %>%
  st_set_crs(4326)
ica_scores <- ica$S %>% 
  as_tibble() %>%
  bind_cols(modern_sites, .) %>%
  select(V1:V9) %>%
  st_set_crs(4326)
```

```{r echo=FALSE}
pc_scores %>%
  as_tibble() %>%
  pivot_longer(PC1:PC9) %>%
  arrange(abs(value)) %>%
  st_as_sf() %>%
ggplot() +
  geom_sf(aes(color = value)) + 
  facet_wrap(~name) +
  geom_sf(data = china %>% select(-name), fill= NA) +
  scale_color_distiller(palette = 'RdBu', limits = c(-88, 88), direction = -1) +
  theme_bw()
```
```{r echo=FALSE}
pc_scores_s %>%
  as_tibble() %>%
  pivot_longer(PC1:PC9) %>%
  arrange(abs(value)) %>%
  st_as_sf() %>% 
ggplot() +
  geom_sf(aes(color = value)) + 
  facet_wrap(~name) +
  geom_sf(data = china %>% select(-name), fill= NA) +
  scale_color_distiller(palette = 'RdBu', limits = c(-10, 10), direction = -1) +
  theme_bw()
```
```{r echo=FALSE}
ica_scores %>%
  as_tibble() %>%
  pivot_longer(V1:V9) %>%
  arrange(abs(value)) %>%
  st_as_sf() %>% 
ggplot() +
  geom_sf(aes(color = value)) + 
  facet_wrap(~name) +
  geom_sf(data = china %>% select(-name), fill= NA) +
  scale_color_distiller(palette = 'RdBu', limits = c(-11, 11), direction = -1) +
  theme_bw()
```

```{r}
ggplot() +
    geom_sf(data= china) +
  geom_sf(data = pc_scores, aes(color = PC1)) + 
  scale_color_scico(palette = 'bamako', limits = c(-70, 70), direction = -1) +
  theme_bw() +
  ggtitle('Modern Arboreal/Non-Arboreal Pollen concentration')
```

```{r}
ggplot() +
    geom_sf(data= china) +
  geom_sf(data = pc_scores, aes(color = PC2)) + 
  scale_color_scico(palette = 'bamako', limits = c(-80, 80), direction = -1) +
  theme_bw() +
  ggtitle('Modern Sedge/Non-Sedge Pollen concentration')
```

## Climate data

```{r}
cmfd <- list.files('~/Downloads/CMFD', full.names = TRUE) %>% map(~read_ncdf(., eps = 0.001, var = 'temp') %>%
  mutate(temp = units::set_units(temp, degree_C) %>% units::drop_units()) %>%
  transmute(gdd0 = if_else(temp > 0, temp, 0) - 0,
            gdd5 = if_else(temp > 5.5, temp, 5.5) - 5.5) %>%
  st_apply(1:2, sum) %>%
  merge()) %>%
  do.call(c, .)%>%
  merge(name = 'time') %>%
  st_set_dimensions('time', values = 1979:2018) %>%
  split('attributes')
gc()
```

```{r}
ggplot() +
  geom_stars(data = st_apply(cmfd['gdd0'], 1:2, mean)) +
  scale_fill_viridis_c(option = 'magma', na.value = NA) +
  theme_bw()
ggplot() +
  geom_stars(data = st_apply(cmfd['gdd5'], 1:2, mean)) +
  scale_fill_viridis_c(option = 'magma', na.value = NA) +
  theme_bw()
```
```{r}
pts <- st_extract(st_apply(cmfd, 1:2, mean), at = modern_sites)
```

Remove sites that fall outside the CMFD boundaries
```{r}
outside_ids <- pts %>%
  bind_cols(modern_sites) %>%
  filter(is.na(gdd0)) %>%
  pull(id)

modern_sites <- modern_sites %>% 
  filter(!(id %in% outside_ids))

modern_pollen <- modern_pollen %>%
  rownames_to_column() %>%
  filter(!(as.numeric(rowname) %in% outside_ids))

ggplot() +
  geom_stars(data = cmfd[1,,,1]) +
  geom_sf(data = pts, aes(color = is.na(gdd0))) +
  scale_color_viridis_d()
```

```{r}
pts <- st_extract(st_apply(cmfd, 1:2, mean), at = modern_sites)
```


```{r}
plot(m3)
```

```{r}
plot(m4)
```
Only by incorporating space do you get a good fit above r2 of .50 and an RMSE competitive with MAT.
```{r}
space_mod <- bam(gdd10 ~ s(PC1, bs = 'cr') + s(PC2, bs = 'cr') + s(PC3, bs = 'cr') + s(PC4, bs = 'cr') + s(PC5, bs = 'cr') + te(x, y), select = TRUE, data = dat_gam)

nospace_mod <- bam(gdd10 ~ s(PC1, bs = 'cr') + s(PC2, bs = 'cr') + s(PC3, bs = 'cr') + s(PC4, bs = 'cr') + s(PC5, bs = 'cr'), select = TRUE, data = dat_gam)

gam.check(space_mod);gam.check(nospace_mod)
summary(space_mod);summary(nospace_mod)
plot(space_mod, residuals = TRUE, scheme = 2);plot(nospace_mod, residuals = TRUE, scheme = 2)
residuals(space_mod)^2 %>% mean %>% sqrt;residuals(nospace_mod)^2 %>% mean %>% sqrt

  AIC(nospace_mod,space_mod)
```

```{r}
pc_scores %>%
  ggplot(aes(PC1, PC2)) +
  geom_point(alpha = .2)
```
How about the forward model?
```{r}
m5 <- gam(PC1 ~ s(mat, bs = 'cr') + s(npp, bs = 'cr') + s(tap, bs = 'cr') + s(gdd10, bs = 'cr'), data = dat, method = 'REML', select = TRUE)

summary(m5)$r.sq
```
```{r}
plot(m5, scale = 0)
```

## Modern Analogue Technique

```{r}
dat <- analogue::join(modern_pollen, fossil_pollen, verbose = TRUE, type = 'inner')
dat2 <- analogue::join(modern_pollen, fossil_pollen2, verbose = TRUE, type = 'inner')
dat3 <- analogue::join(modern_pollen, fossil_pollen3, verbose = TRUE, type = 'inner')
```

H-block cross validation

```{r}
dist_mat <- st_distance(pts)

mat_gdd0_hblock <- MAT(dat$modern_pollen / 100, as_tibble(pts)$gdd0 , lean = FALSE) %>%
  crossval(cv.method = 'h-block', h.cutoff = units::set_units(100000, m), h.dist = dist_mat)
mat_gdd5 <- MAT(dat$modern_pollen / 100, as_tibble(pts)$gdd5, lean = FALSE) %>%
  crossval(cv.method = 'lgo', ngroups = 5)

screeplot(mat_gdd0);screeplot(mat_gdd0_hblock)

```



# Past Climate

Now reconstruct past climates at the fossil pollen sample sites.

```{r}
elev <- raster::getData('alt', country = 'CHN')

tibet_hex <- tibet %>%
  st_make_grid(n = 100, square = FALSE) %>% 
  st_as_sf() %>% 
  rename(geometry = x) %>%
  mutate(x = st_coordinates(st_centroid(.))[,1],
         y = st_coordinates(st_centroid(.))[,2]) %>%
  filter(st_within(geometry, tibet, sparse = FALSE)) %>%
  mutate(id = 1:n(),
         Alt = exact_extract(elev, ., 'mean'))
```


```{r}
ggplot() +
  geom_sf(data = tibet) +
  geom_sf(data = filter(fossil_sites, 
                        st_is_within_distance(geometry, tibet, dist = 400000, sparse = FALSE)), size = 2) +
  geom_sf(data = filter(fossil_sites3, 
                        st_is_within_distance(geometry, tibet, dist = 400000, sparse = FALSE)), color = 'green') 
```


```{r}
pastgdd0 <- predict(mat_gdd0, dat$fossil_pollen / 100, k = 3)
pastgdd5 <- predict(mat_gdd5, dat$fossil_pollen / 100, k = 3)
pastgdd10 <- predict(mat_gdd10, dat$fossil_pollen / 100, k = 3)

tibet_recon <- fossil_sites %>%
  mutate(gdd0 = pastgdd0$fit[,2],
         gdd5 = pastgdd5$fit[,2],
         gdd10 = pastgdd10$fit[,2],
         x = st_coordinates(geometry)[,1],
         y = st_coordinates(geometry)[,2]) %>%
  filter(Age <= 7000) %>%
  filter(st_is_within_distance(geometry, tibet, dist = 400000, sparse = FALSE)) %>% # only get sites within 100km of the plateau
  st_drop_geometry() %>%
  as_tibble()
```

## Interpolation

```{r}

test <- bam(gdd0 ~ te(x, y, Age, d = c(2, 1), bs = c('tp', 'cr')), data = tibet_recon)
test2 <- bam(gdd0 ~ te(x, y, Age, d = c(2, 1), bs = c('tp', 'cr')) + s(Alt, k = 5, bs = 'cr'), data = tibet_recon)
test3 <- bam(gdd0 ~ te(x, y, Age, k = c(50, 20), d = c(2, 1), bs = c('tp', 'cr')), data = tibet_recon)
test4 <- bam(gdd0 ~ te(x, y, Age, k = c(50, 20), d = c(2, 1), bs = c('tp', 'cr')) + s(Alt, k = 5, bs = 'cr'), data = tibet_recon)

#

summary(test);summary(test2);summary(test3);summary(test4)
gam.check(test2)
plot(test);plot(test2);plot(test3);plot(test4)
```

```{r}
library(fields)
test_tps <- Tps(tibet_recon[,c(7,8,2,3)], tibet_recon$gdd0)
plot(test_tps)
summary(test_tps)
```
```{r}
test_tps2 <- Tps(tibet_recon[,c(7,8,3)], tibet_recon$gdd0, Z = tibet_recon$Alt)
plot(test_tps2)
summary(test_tps)
```
```{r}
test_tps3 <- Tps(tibet_recon[,c(7,8,3)], tibet_recon$gdd0)
plot(test_tps3)
```


```{r}
map_dfr(c(0, 6000), ~ mutate(tibet_hex, Age = .)) %>%
    mutate(pred = predict(test4, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  ggplot() + 
  geom_sf(aes(fill = pred) ,size = .1) + 
  scale_fill_viridis_c(option = 'magma', limits = c(0, 4000)) +
  facet_wrap(~Age) + 
  geom_sf(data = tibet, fill = NA) +
  geom_sf(data = fossil_sites %>% filter(Age <= 9000) %>% group_by(site) %>% slice_sample(n = 1) %>% select(-Age) %>% ungroup %>%
  filter(st_is_within_distance(geometry, tibet, dist = 400000, sparse = FALSE)), alpha = .5) +
  theme_bw() 
```

```{r}
new_prd <- map_dfr(c(0), ~ mutate(tibet_hex, Age = .)) %>%
    as_tibble %>%
  select(x,y,Alt, Age) %>%
remove_missing() %>%
      mutate(pred = predict(test_tps,x = .)) %>%
  mutate(pred = if_else(pred < 0, 0, pred))
map_dfr(c(0), ~ mutate(tibet_hex, Age = .)) %>%
  remove_missing() %>%
  mutate(pred = new_prd$pred) %>%
  ggplot() + 
  geom_sf(aes(fill = pred) ,size = .1) + 
  scale_fill_viridis_c(option = 'magma') +
  facet_wrap(~Age) + 
  geom_sf(data = tibet, fill = NA) +
  geom_sf(data = fossil_sites %>% filter(Age <= 9000) %>% group_by(site) %>% slice_sample(n = 1) %>% select(-Age) %>% ungroup %>%
  filter(st_is_within_distance(geometry, tibet, dist = 400000, sparse = FALSE)), alpha = .5) +
  theme_bw() 
```
```{r}
new_prd2 <- map_dfr(c(0), ~ mutate(tibet_hex, Age = .)) %>%
    as_tibble %>%
  select(x,y,Alt, Age) %>%
remove_missing() %>%
      mutate(pred = predict(test_tps2,x = .[,c(1,2,4)], Z = .$Alt)) %>%
  mutate(pred = if_else(pred < 0, 0, pred))
map_dfr(c(0), ~ mutate(tibet_hex, Age = .)) %>%
  remove_missing() %>%
  mutate(pred = new_prd2$pred) %>%
  ggplot() + 
  geom_sf(aes(fill = pred) ,size = .1) + 
  scale_fill_viridis_c(option = 'magma') +
  facet_wrap(~Age) + 
  geom_sf(data = tibet, fill = NA) +
  geom_sf(data = fossil_sites %>% filter(Age <= 9000) %>% group_by(site) %>% slice_sample(n = 1) %>% select(-Age) %>% ungroup %>%
  filter(st_is_within_distance(geometry, tibet, dist = 400000, sparse = FALSE)), alpha = .5) +
  theme_bw() 
```
```{r}
new_prd3 <- map_dfr(c(0), ~ mutate(tibet_hex, Age = .)) %>%
    as_tibble %>%
  select(x,y,Alt, Age) %>%
remove_missing() %>%
      mutate(pred = predict(test_tps3,x = .[,c(1,2,4)])) %>%
  mutate(pred = if_else(pred < 0, 0, pred))
map_dfr(c(0), ~ mutate(tibet_hex, Age = .)) %>%
  remove_missing() %>%
  mutate(pred = new_prd3$pred) %>%
  ggplot() + 
  geom_sf(aes(fill = pred) ,size = .1) + 
  scale_fill_viridis_c(option = 'magma') +
  facet_wrap(~Age) + 
  geom_sf(data = tibet, fill = NA) +
  geom_sf(data = fossil_sites %>% filter(Age <= 9000) %>% group_by(site) %>% slice_sample(n = 1) %>% select(-Age) %>% ungroup %>%
  filter(st_is_within_distance(geometry, tibet, dist = 400000, sparse = FALSE)), alpha = .5) +
  theme_bw() 
```

```{r}
tibet_hex %>% transmute(gdd0 = exact_extract(as(clim['GDD0'], 'Raster'), ., 'mean')) %>% ggplot() + geom_sf(aes(fill = gdd0), size = .1) + scale_fill_viridis_c(option = 'magma') + theme_bw()
```
The base gam doesn't to a great job of interpolating
```{r}
tibet_hex %>% mutate(Age = -50) %>%
  mutate(gdd0 = exact_extract(as(clim['GDD0'], 'Raster'), ., 'mean'),
  pred = predict(test4, ., type = 'response'),
  pred = if_else(pred < 0, 0, pred)) %>%
  ggplot(aes(gdd0, pred)) +
  geom_point() +
  geom_smooth() +
  theme_bw()
```
```{r}
tibet_hex %>%
  mutate(gdd0 = exact_extract(as(clim['GDD0'], 'Raster'), ., 'mean')) %>%
  ggplot(aes(Alt, gdd0)) +
  geom_point() +
  geom_smooth() +
  theme_bw()
```



```{r}
mutate(china_hex, Age = 0) %>%
    mutate(pred = predict(test, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  mutate(Age = 6000) %>%
      mutate(pred2 = predict(test, ., type = 'response')) %>%
  mutate(pred2 = if_else(pred2 < 0, 0, pred2)) %>%
  mutate(pred = pred - pred2) %>%
  ggplot() + geom_sf(aes(fill = pred) ,size = .1) + scale_fill_distiller(palette ='RdBu', limits = c(-1900, 1900)) +
  ggtitle('Change in Growing Degree Days', 'Mid Holocene to Present') +
  theme_bw() 
```

```{r}
mutate(china_hex, Age = 3500) %>%
    mutate(pred = predict(test, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  mutate(Age = 6000) %>%
      mutate(pred2 = predict(test, ., type = 'response')) %>%
  mutate(pred2 = if_else(pred2 < 0, 0, pred2)) %>%
  mutate(pred = pred - pred2) %>%
  ggplot() + geom_sf(aes(fill = pred) ,size = .1) + scale_fill_distiller(palette ='RdBu') +
  ggtitle('Change in Growing Degree Days', 'Mid Holocene to Present') +
  geom_point(data = fossil_loc %>% filter(Long > 60), aes(Long, Lat)) +
  theme_bw() 
```

```{r}
map_dfr(seq(7000, 0, -1000), ~ mutate(tibet_hex, Age = .)) %>%
    mutate(pred = predict(test4, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  ggplot() + geom_sf(aes(fill = pred > 1907) ,size = .1) + scale_fill_viridis_d(option = 'magma') +
    facet_wrap(~Age) +
  theme_bw() 
```

## Bins
```{r}
library(exactextractr)
gdd_dgg <- exact_extract(as(clim['GDD0'], 'Raster'), tibet_hex, 'mean')

china_hex %>% transmute(gdd_dgg) %>% ggplot() + geom_sf(aes(fill = gdd_dgg), size = .1) + scale_fill_viridis_c(option = 'magma') + theme_bw()
```
```{r}
china_hex %>% transmute(gdd_dgg) %>% ggplot() + geom_sf(aes(fill = gdd_dgg > 2900), size = .1) + scale_fill_viridis_d(option = 'magma') + theme_bw()
```
repeat with bins
```{r}
read_excel('china pollen data/China modern pollen - data.xlsx', sheet = 2, skip = 1) %>%
  slice(-(1:2)) %>%
  st_as_sf(coords = c(4,5), crs = 4326) %>%
 st_intersects(china_hex, .) %>%
  lengths %>%
  transmute(china_hex, .) %>% plot
```

```{r}
mod_hex_pol <- read_excel('china pollen data/China modern pollen - data.xlsx', sheet = 2, skip = 1) %>%
  slice(-(1:2)) %>%
  st_as_sf(coords = c(4,5), crs = 4326) %>%
  st_join(china_hex, join = st_within) %>% 
  select(Abelia:id) %>%
  st_drop_geometry() %>%
  mutate(across(everything(), as.numeric)) %>%
    group_by(id) %>%
  summarise(across(everything(), mean)) %>%
  left_join(select(china_hex, -Long, -Lat), .) %>%
  select(-Long, -Lat) %>% 
  st_drop_geometry() %>% as_tibble%>% select(-id) %>% 
  `/`(100)  



dat2 <- mod_hex_pol %>% mutate(gdd  = gdd_dgg) %>% remove_missing()

mat_bin <- rioja::MAT(dat2 %>% select(-gdd), dat2$gdd)
screeplot(mat_bin)
```


```{r}

fossil_loc_bin <- read_excel('china pollen data/China west fossil pollen - data.xlsx') %>%
  slice(-(1:2)) %>%
  st_as_sf(coords = c(2,3), crs = 4326) %>%
  st_join(china_hex, join = st_within) %>% 
  select(Age,`Abies+Picea`:id) %>%
  st_drop_geometry() %>%
  mutate(across(everything(), as.numeric)) %>%
  left_join(select(china_hex, -Long, -Lat), .) %>%
  select(-Long, -Lat)


fossil_poll_bin <- fossil_loc_bin %>% 
  st_drop_geometry() %>% as_tibble %>% remove_missing() %>% group_by(id, Age) %>% summarise(across(everything(), mean), .groups = 'drop') %>% mutate(no = interaction(id, Age, sep = '_')) %>% column_to_rownames(var = 'no') %>% 
  select(-id, -Age) %>%
  `/`(100)  

fossil_poll_bin_ages <- fossil_loc_bin %>% 
  st_drop_geometry() %>% as_tibble%>% select(id, Age) %>% remove_missing()

past_gdd_bin <- predict(mat_bin, fossil_poll_bin, k = 4)

recon_mod <- as_tibble(past_gdd_bin$fit, rownames = 'id') %>%
  separate(id, into = c('id', 'Age'), sep = '_') %>% 
  mutate(id = as.numeric(id), Age = as.numeric(Age)) %>%
 # left_join(fossil_poll_bin_ages) %>%
  select(id, Age, gdd10 = MAT.wm) %>%
  left_join(china_hex, by = 'id') %>%
  filter(Long > 60, Age < 10000) %>%
   bam(gdd10 ~ te(Long, Lat, Age, k = c(80, 20), d = c(2, 1), bs = c('tp', 'cr')), data = .)

summary(recon_mod)
plot(recon_mod)
```
```{r}

# not binned until end
mutate(china_hex, Age = 0) %>%
    mutate(pred = predict(test, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  mutate(Age = 6000) %>%
      mutate(pred2 = predict(test, ., type = 'response')) %>%
  mutate(pred2 = if_else(pred2 < 0, 0, pred2)) %>%
  mutate(pred = pred - pred2) %>%
  ggplot() + geom_sf(aes(fill = pred) ,size = .1) + scale_fill_distiller(palette ='RdBu', limits = c(-1900, 1900)) +
  theme_bw() 

#binned
mutate(china_hex, Age = 0) %>%
    mutate(pred = predict(recon_mod, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  mutate(Age = 6000) %>%
      mutate(pred2 = predict(recon_mod, ., type = 'response')) %>%
  mutate(pred2 = if_else(pred2 < 0, 0, pred2)) %>%
  mutate(pred = pred - pred2) %>%
  ggplot() + geom_sf(aes(fill = pred) ,size = .1) + scale_fill_distiller(palette ='RdBu', limits = c(-1900, 1900)) +
  theme_bw() 
```
```{r}
#why isn't there more variability here?
as_tibble(pastgdd$fit, rownames = 'id') %>%
  separate(id, into = c('id', 'Age')) %>%
  mutate( Age = as.numeric(Age)) %>%
 # left_join(fossil_poll_bin_ages) %>%
  select(id, Age, gdd10 = MAT.wm) %>%
  filter(Age < 7500) %>%
  ggplot(aes(-Age, gdd10, color = (id))) + 
  geom_point(alpha = .5) +
  geom_line(alpha = .3) +
  theme_bw() +
  guides(color = 'none')
```

```{r}
data(columb.polys) ## district shapes list

spdep::poly2nb(china_hex )
columb.polys
```

```{r}
raster::getData()
```

## FRK

```{r}
library(FRK)
library(sp)
data("AIRS_05_2003")
AIRS_05_2003
class(AIRS_05_2003)
coordinates(AIRS_05_2003) <- ~lon + lat
proj4string(AIRS_05_2003) <- CRS("+proj=longlat +ellps=sphere")
isea3h_sp_poldf <- auto_BAUs(manifold = sphere(), isea3h_res = 6, type = "hex", data = AIRS_05_2003)
isea3h_sp_poldf$fs <- 1
plot(isea3h_sp_poldf)
```

```{r}
G <- auto_basis(manifold = sphere(), data = AIRS_05_2003_t, nres = 3,isea3h_lo = 2, type = "bisquare")
```

```{r}
  mutate(china_hex, Age = 6000) %>%
      mutate(pred = predict(test, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  ggplot() + geom_sf(aes(fill = pred) ,size = .1) + scale_fill_distiller(palette ='RdBu') +
  theme_bw() 
```



```{r}
mh_test_mod <- recon %>%
  filter(Age > 5550 & Age < 6450 ) %>%
  bam(gdd10 ~ te(Long, Lat, k = 10), data = .)
mutate(china_hex) %>%
    mutate(pred = predict(mh_test_mod, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  ggplot() + geom_sf(aes(fill = pred) ,size = .1) + scale_fill_distiller(palette ='RdBu') +
  theme_bw() 
```

```{r}
mutate(china_hex, Age = 4200) %>%
    mutate(pred = predict(recon_mod, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  ggplot() + geom_sf(aes(fill = pred) ,size = .1) + scale_fill_distiller(palette ='RdBu') +
  theme_bw() 
```

```{r}
  mutate(china_hex, Age = 4200) %>%
      mutate(pred = predict(test, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  ggplot() + geom_sf(aes(fill = pred) ,size = .1) + scale_fill_distiller(palette ='RdBu') +
  theme_bw() 

mutate(china_hex, Age = 0) %>%
    mutate(pred = predict(test, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  mutate(Age = 4200) %>%
      mutate(pred2 = predict(test, ., type = 'response')) %>%
  mutate(pred2 = if_else(pred2 < 0, 0, pred2)) %>%
  mutate(pred = pred - pred2) %>%
  ggplot() + geom_sf(aes(fill = pred) ,size = .1) + scale_fill_distiller(palette ='RdBu') +
  theme_bw() 
```


## Tibet

```{r}
tibet_bbox <- st_bbox(c(xmin = 65, xmax = 105, ymin = 25, ymax = 45), crs = 4326)
tibet_mod <- recon %>%
  filter(between(Lat, 25, 45), between(Long, 65, 105)) %>%
   bam(gdd10 ~ te(Long, Lat, Age, k = c(80, 20), d = c(2, 1), bs = c('tp', 'cr')), data = .)
summary(tibet_mod)

```
```{r}
tibet_hex <- tibet_bbox %>% 
  st_make_grid(n = 50, square = FALSE) %>% 
  st_as_sf() %>% 
  mutate(Long = st_coordinates(st_centroid(.))[,1],
         Lat = st_coordinates(st_centroid(.))[,2]) %>%
  #  filter(st_intersects(x, china, sparse = FALSE)) %>%
  mutate(id = 1:n())

map_dfr(c(0, 2000, 4000, 6000), ~ mutate(tibet_hex, Age = .)) %>%
    mutate(pred = predict(tibet_mod, ., type = 'response')) %>%
  mutate(pred = if_else(pred < 0, 0, pred)) %>%
  ggplot() + geom_sf(aes(fill = pred) ,size = .1) + scale_fill_viridis_c(option = 'magma') +
    facet_wrap(~Age) +
  theme_bw()  +
  geom_point(data = recon %>% filter(between(Lat, 25, 45), between(Long, 65, 105)) %>% group_by(Lat, Long) %>% summarise(Lat = mean(Lat), Long = mean(Long), age = 0), aes(Long, Lat))
```

# Other Methods

Comparison of other pollen reconstruction methods.

```{r}
gdd_wapls <- WAPLS(sqrt(dat$modern_pollen[,-298] / 100), as_tibble(pts)$gdd0, 
                           npls = 10,
                           standx = FALSE) %>% 
  crossval(cv.method = 'h-block', h.cutoff = units::set_units(200000, m), h.dist = dist_mat)
screeplot(gdd_wapls)
plot(gdd_wapls, npls = 4)
```

```{r}
gdd_LWR <- rioja::LWR(sqrt(dat$modern_pollen / 100), as_tibble(pts)$gdd0) %>%
  crossval(cv.method = 'lgo', ngroups = 5)
plot(gdd_LWR)

length(gdd_LWR$x)
plot(gdd_LWR$x, gdd_LWR$predicted[,1])

colSums(dat$modern_pollen)
```

```{r}
gdd_IKFA <- rioja::IKFA(dat$modern_pollen[,-298] / 100, as_tibble(pts)$gdd0, nFact = 15) %>%
  crossval(cv.method = 'h-block', h.cutoff = units::set_units(200000, m), h.dist = dist_mat)

screeplot(gdd_IKFA)
plot(gdd_IKFA, nFact)

gdd_IKFA$ccoef
```

